Declarative Deployment Reconfiguration of Pulverised Collective Adaptive Systems at Runtime \thanks{ This work has been supported by the Italian PRIN Project COMMON-WEARS (2020HCWWLP) and the EU/MUR FSE REACT-EU PON R\&I 2014-2020. }

### 2023-05-21 paper-2023-acsos-pulverization.tex
Renaming DSLs + some fixes @danysk @nicolasfara

@@

From the previous section,
we can identify two main pieces of information for the definition of a deployment plan for a pulverised system:
\begin{enumerate*}[label=\it{(\roman*)}]
  \item the definition of the \emph{target} deployment domain (\emph{infrastructure}), i.e., the specification of the \emph{requirements} posed on hosts in terms of required (or preferred) capabilities; and
  \item the definition of the \emph{deployment plan}, i.e., the allocation of the pulverised components to the infrastructural elements, together with constraints and reconfiguration policies.
\end{enumerate*}
%
Coherently with the \emph{single responsibility principle},
we capture these two descriptive needs into two distinct \acp{dsl}.
%
An overall view of the approach in terms of the provided tools and the main modelling concepts is provided in \Cref{fig:approach}.

@@

\subsection{Infrastructure and Requirements}

The infrastructure \ac{dsl} captures the concepts of
\emph{device type}, \emph{component}, \emph{capability}, and \emph{requirement}.
%
A reference sample is presented in \Cref{lst:system-configuration}.
%
The DSL provides simple means to define capabilities as types in the host language (Lines~\ref{capab:start}--\ref{capab:end}),
and then to associate them to pulverised components for each device type (Lines~\ref{devcap:start}--\ref{devcap:end}).
%
This \ac{dsl} is meant to define constraints that the application imposes on the infrastructure,
ruling out configurations that are not capable to support the system
(for instance, deploying the sensor component of a device in a host that exposes no sensor).
%
All combinations of components and capabilities that are defined as supported become instead amenable as \emph{targets} of deployment or reconfiguration---forming what we may call the \emph{deployment range}.
%\emph{runtime reconfiguration}.
%
In the example, e.g., component \texttt{Behavior} in \texttt{iot-sensor} is defined as executable on any of host exposing
\texttt{HighCPU} and/or \texttt{EmbeddedDevice} as capabilities (\Cref{devcap:b}):
it implies that the component can be dynamically moved to any host offering any such capability.

@@

\begin{lstlisting}[
    float=ht,
    language=pulv,
    caption={
        Example infrastructure configuration,
        defining three capabilities and a system composed of two device types,
        each component of which is then bound to the set of capabilities it needs to execute.
    },
    label=lst:system-configuration,
    escapechar=\%,
    numbers=left
]
object HighCPU : Capability%\label{capab:start}%
object LowLatencyComm : Capability
object EmbeddedDevice : Capability%\label{capab:end}%

@@

\subsection{Deployment and Reconfiguration}

The second \ac{dsl} supports the definition of the mapping between pulverised components and specific hosts,
as well as the definition of reactive reconfiguration policies.
%
It works with the concepts of \texttt{Host}
(associated to \texttt{Capability}),
\texttt{ReconfigurationEvent},
and \texttt{ReconfigurationRule}.
%
Using \Cref{lst:runtime-configuration} as a reference example,
we show how reconfiguration events (Lines~\ref{code:reconfigev:start}--\ref{code:reconfigev:end}) can be captured at the type-system level
by expressing them as a predicate over an asynchronous flow of \texttt{events} influencing the system's state.
%
Similarly, hosts can be configured by defining their name and capabilities (Lines~\ref{code:hosts:start}--\ref{code:hosts:end}).
%
Once the hosts composing the system have been configured,
the entire infrastructure can be set up (Lines~\ref{code:infra:start}--\ref{code:infra:end}) by defining,
for each logical device,
which host should host each pulverised component (Lines~\ref{code:mapping:start}--\ref{code:mapping:end}).
%
Additionally,
reconfiguration rules (Lines~\ref{code:reconfig:start}--\ref{code:reconfig:end}) can be defined to specify how the system should react to reconfiguration events.
%
In this context, the \ac{dsl} exposes a special syntax to express the migration of a pulverised component on a different host.
%
Crucially, being the DSL internal,
specialised behaviour not supported by the \ac{dsl} can be defined in the host language directly:
configuration blocks surrounded by curly brackets are,
in fact,
plain lambda expressions, supporting any kind of computation.

>>>>

\newcommand{\dslSys}[0]{\emph{System \ac{dsl}}}
\newcommand{\dslDep}[0]{\emph{Deployment \ac{dsl}}}

@@

From the previous section,
we can identify two main pieces of information for the definition of a deployment plan for a pulverised system:
\begin{enumerate*}[label=\it{(\roman*)}]
  \item the definition of the \emph{target} deployment domain (\emph{infrastructure}), i.e., the specification of the \emph{requirements} posed on hosts in terms of required (or preferred) capabilities; and
  \item the definition of the \emph{deployment plan}, i.e., the allocation of the pulverised components to the infrastructural elements, together with constraints and reconfiguration policies.
\end{enumerate*}
%
Coherently with the \emph{single responsibility principle},
we capture these two descriptive needs into two distinct \acp{dsl}: the \dslSys{} and the \dslDep{}.
%
An overall view of the approach in terms of the provided tools and the main modelling concepts is provided in \Cref{fig:approach}.

@@

\subsection{The \dslSys{}: Components and Required Capabilities}

The \dslSys{} captures the concepts of
\emph{device type}, \emph{component}, \emph{capability}, and \emph{requirement}.
%
A reference sample is presented in \Cref{lst:system-configuration}.
%
The DSL provides simple means to define capabilities as types in the host language (Lines~\ref{capab:start}--\ref{capab:end}),
and then to associate them to pulverised components for each device type (Lines~\ref{devcap:start}--\ref{devcap:end}).
%
This \ac{dsl} is meant to define constraints that the application imposes on the infrastructure,
ruling out configurations that are not capable to support the system
(for instance, deploying the sensor component of a device in a host that exposes no sensor).
%
All combinations of components and capabilities that are defined as supported become instead amenable as \emph{targets} of deployment or reconfiguration---forming what we may call the \emph{deployment range}.
%\emph{runtime reconfiguration}.
%
In the example, e.g., component \texttt{Behavior} in \texttt{iot-sensor} is defined as executable on any of host exposing
\texttt{HighCPU} and/or \texttt{EmbeddedDevice} as capabilities (\Cref{devcap:b}):
it implies that the component can be dynamically moved to any host offering any such capability.

@@

\begin{lstlisting}[
    float=ht,
    language=pulv,
    caption={
        Example of \dslSys{} usage.
        This code snippet defines three capabilities and a logical system composed of two device types,
        whose components are bound to the set of capabilities they needs to execute.
    },
    label=lst:system-configuration,
    escapechar=\%,
    numbers=left
]
object HighCPU : Capability%\label{capab:start}%
object LowLatencyComm : Capability
object EmbeddedDevice : Capability%\label{capab:end}%

@@

\subsection{The \dslDep{}: Deployment Domain, Mapping, and Reconfiguration}

The \dslDep{} supports the definition of the mapping between pulverised components and specific hosts,
as well as the definition of reactive reconfiguration policies.
%
It works with the concepts of \texttt{Host}
(associated to \texttt{Capability}),
\texttt{ReconfigurationEvent},
and \texttt{ReconfigurationRule}.
%
Using \Cref{lst:runtime-configuration} as a reference example,
we show how reconfiguration events (Lines~\ref{code:reconfigev:start}--\ref{code:reconfigev:end}) can be captured at the type-system level
by expressing them as a predicate over an asynchronous flow of \texttt{events} influencing the system's state.
%
Similarly, hosts can be configured by defining their name and capabilities (Lines~\ref{code:hosts:start}--\ref{code:hosts:end}).
%
Once the hosts composing the system have been configured,
the entire infrastructure can be set up (Lines~\ref{code:infra:start}--\ref{code:infra:end}) by defining,
for each logical device,
which host should host each pulverised component (Lines~\ref{code:mapping:start}--\ref{code:mapping:end}).
%
Additionally,
reconfiguration rules (Lines~\ref{code:reconfig:start}--\ref{code:reconfig:end}) can be defined to specify how the system should react to reconfiguration events.
%
In this context, the \ac{dsl} exposes a special syntax to express the migration of a pulverised component on a different host.
%
Crucially, being the DSL internal,
specialised behaviour not supported by the \ac{dsl} can be defined in the host language directly:
configuration blocks surrounded by curly brackets are,
in fact,
plain lambda expressions, supporting any kind of computation.
